---
title: "Section 4: Cross-sectional comparison of serum and nail measures, coefficient of variation. RTI sample."
output:
  pdf_document:
    number_sections: yes
    toc: no
    highlight: tango
  html_document:
    theme: united
    toc: no
urlcolor: blue
editor_options:
  chunk_output_type: console
bibliography: ../iron.bib
---

# Cross-sectional comparison of serum and nail coefficient of variation with methods from R cvequality package. # {#cv}

```{r, include=FALSE}
knitr::opts_chunk$set(echo = F,
                      results = 'hide',
                      warning = F,
                      fig.pos = 'H')

#runif(1, 0, 10^8)
set.seed(74268794)

```


```{r, include=F}
# bring in packages
library(cvequality)
library(dplyr)
library(tidyr)
library(plyr)
library(kableExtra)
library(ggplot2)
library(data.table)

```


```{r, results='hide', echo=F}

load(file="../sections/updated-data-iron.RData") # sub.dat is subset, all.dat is total data

nail.names2
iron.names2
serum.names2


load(file="RTI-dat.RData") # all.dat2 from section2-RTI.Rmd
```

```{r}
# Get ids of people with paired measures for both time points
nrow(all.dat2)
dt = data.table(all.dat2)

# Counts of non-missing combinations of fe/fertn/fesat + NAIL values at baseline 
misst1 = dcast(dt, (!(is.na(UMN_Iron_Baseline_FE) | is.na(UMN_Iron_Baseline_FERTN) | is.na(UMN_Iron_Baseline_FESAT))) ~ (!(is.na(RTI_Toenail_fe_Adj))),
      value.var='PSID', length)
misst1

# Counts of non-missing combinations of fe values at baseline and second measure
misst2 = dcast(dt, ( (!(is.na(all.dat2$UMN_Iron_SCL_FE) | is.na(all.dat2$UMN_Iron_SCL_FERTN) | is.na(all.dat2$UMN_Iron_SCL_FESAT))) ) ~ (!(is.na(DC_SCL_Toenail_Fe))),
      value.var='PSID', length)
misst2

paired.id.t1 = all.dat2[(!(is.na(all.dat2$UMN_Iron_Baseline_FE) | is.na(all.dat2$UMN_Iron_Baseline_FERTN) | is.na(all.dat2$UMN_Iron_Baseline_FESAT))) & (!(is.na(all.dat2$RTI_Toenail_fe_Adj))), ]$PSID
length(paired.id.t1)


paired.id.t2 = all.dat2[(!(is.na(all.dat2$UMN_Iron_SCL_FE) | is.na(all.dat2$UMN_Iron_SCL_FERTN) | is.na(all.dat2$UMN_Iron_SCL_FESAT))) & (!(is.na(all.dat2$DC_SCL_Toenail_Fe))), ]$PSID
length(paired.id.t2)

summary(all.dat2[all.dat2$PSID %in% paired.id.t1, c("UMN_Iron_SCL_FE", "DC_SCL_Toenail_Fe", "RTI_Toenail_fe_Adj",
                                                 "UMN_Iron_SCL_FERTN")])
hist(all.dat2$DC_SCL_Toenail_Fe)
hist(all.dat2$RTI_Toenail_fe_Adj)

```


## Coefficient of variation for nail and serum measures at baseline


```{r}
# Make long style data frame for all 3 serum iron measures

dat.long <- 
  all.dat2[all.dat2$PSID %in% paired.id.t1,] %>% 
  dplyr::select(c(UMN_Iron_Baseline_FE, UMN_Iron_Baseline_FERTN, UMN_Iron_Baseline_FESAT,
                  RTI_Toenail_fe_Adj,
                  subcohort.nocases,
                  PSID)) %>% 
  pivot_longer(cols = -c("PSID", 'subcohort.nocases'),
               values_drop_na=T)

levels(factor(dat.long$name))

dat.long = within(dat.long, {
  no.case = factor(subcohort.nocases,
                   labels=c("No", "Yes"))
  name.f = factor(name, labels=c("Toenail Iron", "Serum Iron", "Ferritin", "Transferrin Saturation"))
})

head(dat.long)
dim(dat.long)
length(unique(dat.long$PSID))

# remove outlier from nail data frame
nrow(dat.long)


extreme.id = all.dat2[which(all.dat2$RTI_Toenail_fe_Adj > quantile(all.dat2$RTI_Toenail_fe_Adj, 0.998, na.rm=T)),]$PSID # remove people with extreme fe values 
extreme.id

all.dat2[all.dat2$PSID %in% extreme.id,c("PSID",
                                       "RTI_Toenail_fe_Adj")]

dat.long = dat.long[!(dat.long$PSID %in% extreme.id),]
nrow(dat.long)
summary(dat.long)


```




```{r, results='markup'}

# with outlier removed
cv.f<-function(df){
  x=log(df$value)
  sd(x)/mean(x)
}

# get cv info
cv.1 <- dlply(dat.long, "name.f", cv.f)

# return a data frame
cv1.info=ldply(cv.1)
names(cv1.info) = c("Measure", "CV")
cv1.info

# counts
cv.f.n<-function(df){
  x=df$value
  length(x)
}
# return a data frame with frequency for each cv
cv.1.n = dlply(dat.long, "name.f", cv.f.n)

cv1.info.n = ldply(cv.1.n)
names(cv1.info.n) = c("Measure", "n")
cv1.info.n

```


## Characterize and test equality of CV for the nail and serum measures

Source: https://cran.r-project.org/web/packages/cvequality/vignettes/how_to_test_CVs.html

Note: This test assumes independence between samples. Need to find a method that accounts for dependence and drawn from different distributions.

```{r, results='markup'}

# source: https://cran.r-project.org/web/packages/cvequality/vignettes/how_to_test_CVs.html

# take a quick look
kable(head(dat.long),
      caption = "Preview of first few rows of the data")
```

## Look at variation of variables.

```{r}
ggplot(dat.long,
       aes(name.f,
           value)) +
  geom_boxplot() +
#  geom_quasirandom(alpha = 0.02) +
  theme_bw()
```

## Look at variation of log-transformed variables.

```{r}

p.s4 = ggplot(dat.long,
       aes(name.f,
           log(value))) +
  geom_boxplot() +
#  geom_quasirandom(alpha = 0.02) +
  theme_bw(base_size=20) + 
  labs(x="Measure",
       y="ln(value)")

p.s4
```


## Test if variation in log-transformed nail data is greater than the serum data

```{r}

get.test = function(iron.measure){

  # iron.measure="Serum Iron"
  sub.dat = dat.long[dat.long$name.f %in% c(iron.measure, "Toenail Iron"),]
  x = log(sub.dat$value); head(x)
  y = sub.dat$name.f;  y = factor(y); head(y)

  iron.asymptotic.test = asymptotic_test(x,y)

  iron.mlrt.test = mslr_test(nr = 1e4,
                             log(x), y)
  
  return(data.frame(serum.val = iron.measure,
           D_AD = as.numeric(iron.asymptotic.test$D_AD), 
           p_value = as.numeric(iron.asymptotic.test$p_value),
           MSLRT = as.numeric(iron.mlrt.test$MSLRT),
           mslrt.pvalue = as.numeric(iron.mlrt.test$p_value)))
}


```

```{r}
# get people with serum values and no nail measures and all nail measures
psid.nail = unique(dat.long[dat.long$name.f=="Toenail Iron",]$PSID)
length(psid.nail)

psid.fe = unique(dat.long[dat.long$name.f=="Serum Iron",]$PSID)
length(psid.fe)

both = psid.fe[psid.fe %in% psid.nail]
length(both)

dat.long.ind = rbind.data.frame(dat.long[dat.long$name.f=="Toenail Iron",],
                                dat.long[!(dat.long$name.f %in% "Toenail Iron") & !(dat.long$PSID %in% both),])

head(dat.long.ind)

```

Using the [cvequality package](https://rdrr.io/cran/cvequality/f/vignettes/how_to_test_CVs.Rmd) in R, calculate the 'Asymptotic test for the equality of coefficients of variation from k populations' (Feltz and Miller 1996) and 'Modified signed-likelihood ratio test (SLRT) for equality of CVs' (Krishnamoorthy and Lee 2014) tests.


```{r, results='markup'}

# get all nail measures and no serum values for people with nail values

s4dat = rbind(get.test("Serum Iron"),
      get.test("Ferritin"),
      get.test("Transferrin Saturation"))

s4dat

```


Strong evidence that there are differences in variability between the nail and serum measures.


```{r}

save(s4dat, cv1.info, p.s4, file="section4-RTI-dat-log.RData")

```

